---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

# Set up

Load packages

```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
```

Load custom functions

```{r}
parent_dir <- dirname(getwd()) # have to go up one directory to find the script with all functions
source(paste0(parent_dir,"/51_Uncertainty_all_FUNs.R"))
```

Get all data in one data frame

```{r}
grand_parent_dir <- dirname(parent_dir) # go up two directories to find the data
data_birds_all <- create_df_birds(root_dir = grand_parent_dir)
```

# Calculate Vulnerability

Using default settings (normalization is "minmax", summary method is "sum") 

```{r}
VU_birds <- vulnerabilityFUN(data_birds_all)
head(VU_birds)
```

The default settings calculate the indexes per species/island. If you want just per island, you can set `by_island` to `TRUE`

```{r}
VU_birds_isl <- vulnerabilityFUN(data_birds_all, by_island = TRUE)
head(VU_birds_isl)
```

You can try it with different normalization methods (options are: "minmax", "logminmax" and "rankminmax")

```{r}
VU_birds_logminmax <- vulnerabilityFUN(data_birds_all, norm = "logminmax")
```

Or with different summary methods (options are: "sum", "rank_sum", "TOPSIS", "rank_TOPSIS", "prod", "rank_prod")

```{r}
VU_birds_TOPSIS <- vulnerabilityFUN(data_birds_all, summary_method = "TOPSIS")
```

You can also bootstrap the calculation, to get many replicates, from which you can calculate a median and confidence intervals. Use `n_samples` to set the number of replicates and `prop_samples` to set the proportion of the data each sample uses

```{r}
VU_birds_boot <- 
  vulnerabilityFUN(data_birds_all, 
                   by_island = TRUE,
                   n_samples = 100, 
                   prop_samples = 0.9)
```

Now let's calculate the medians, standard deviation and confidence intervals for each component, using the function `summary_bootstrap`

```{r}
VU_birds_boot_summary <-
summary_bootstrap(VU_birds_boot)

head(VU_birds_boot_summary)
```

The function also has the arguments `lu_cols`, `cc_cols`, `ias_cols`, `sens_cols`, `ac_sp_cols` and `ac_isl_cols`, where you can input the name of the columns containing the data of each component, if they are different from the default. If you are changing the name of the columns, you also have to inform which variables are inversely correlated with their component, using argument `inv_variables`. For example, range size is used to calculate sensitiviy, but the larger the range size, the smaller the sensitivity.

Lastly, you can also calculate each component separately, which will also return the normalized variables used to calculate them:

```{r}
E_birds <- exposureFUN(data_birds_all)
S_birds <- sensitivityFUN(data_birds_all)
AC_birds <- adaptive_capacityFUN(data_birds_all)

head(E_birds)
head(S_birds)
head(AC_birds)
```

# Graphs

Plot distribution of species vulnerability by island

```{r}
arch_isl <- paste0(VU_birds$ARCHIP, VU_birds$Island_name)
island_order <- VU_birds$Island_name[match(sort(arch_isl), arch_isl)] %>% unique # This is to make sure the same islands of each archipelago are together in the graph

VU_birds$Island_name <- factor(VU_birds$Island_name, levels = island_order)

VU_median_mean <- 
  VU_birds %>% 
  group_by(Island_name) %>% 
  summarise(Median = median(VU),
            Mean = mean(VU))

ggplot(VU_birds, aes(VU)) +
  geom_histogram(aes(fill=ARCHIP, col=ARCHIP)) +
  facet_wrap(~Island_name) +  
  geom_segment(data=VU_median_mean,
               aes(x = Median, xend=Median, y=0, yend=10), col='black') +
  geom_segment(data=VU_median_mean,
               aes(x = Mean, xend=Mean, y=0, yend=10), col='blue') +
  theme_bw()
```

Plot median and confidence intervals of vulnerability by island

```{r}
VU_birds_boot_summary$Island_name <- factor(VU_birds_boot_summary$Island_name, levels = island_order)

ggplot(VU_birds_boot_summary, aes(x=median_VU,
                                         xmin=lci_VU,
                                         xmax=uci_VU,
                                         y=Island_name,
                                         fill=ARCHIP, 
                                         col=ARCHIP)) +
  geom_point() +
  geom_errorbar() +
  theme_bw() +
  ggtitle('Sensitivity') +
  xlab('median \u00B1 95% CI') +
  scale_color_manual("Archipelago",
                     values = c("Azores" = brewer.pal(n = 5, name = "Set1")[5],
                                "Canary Islands" = brewer.pal(n = 5, name = "Set1")[2],
                                "Galapagos Islands" = brewer.pal(n = 5, name = "Set1")[3],
                                "Hawaii" = brewer.pal(n = 5, name = "Set1")[1],
                                "Mascarene Islands" = brewer.pal(n = 5, name = "Set1")[4])) +
  scale_fill_manual("Archipelago",
                    values = c("Azores" = brewer.pal(n = 5, name = "Set1")[5],
                               "Canary Islands" = brewer.pal(n = 5, name = "Set1")[2],
                               "Galapagos Islands" = brewer.pal(n = 5, name = "Set1")[3],
                               "Hawaii" = brewer.pal(n = 5, name = "Set1")[1],
                               "Mascarene Islands" = brewer.pal(n = 5, name = "Set1")[4])) +
  theme(legend.position = "bottom") +
  ylab('Island name')
```